---
title: "HW4 - Analysis"
author: "George Dewey"
date: "3/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(jtools)

theme_set(theme_stata())
```

## Analysis

We will run two types of models for this analysis - a linear regression (potentially with optimization) and a neural network-based regression - and compare their performance.

### Load Data

```{r}
ts = readRDS("ts_cohort_analysis")
ts
```

### Linear Regression

Based on the graphics created previously, I am going to included the following predictors in a linear regression model using `hosp_time` - the overall time in hospital measured in days - as the outcome variables:

* Age
* Diagnosis Priority (i.e. `seq_num`)
* Time in ICU (measured in days)
* Survival status
* Gender
* Marital Status
* Ethnicity

I choose not to include time in ER as it is likely highly correlated with time in ICU. While there may be concerns with reverse causation with regarsd to survival status, I am looking at this problem in a purely predictive sense and thus the fact that people certain to die have much shorter hospital stays compared to those who may survive is worth noting.

```{r}
linear1 = lm(hosp_time ~ seq_num + icu_los + factor(hospital_expire_flag) + age + factor(gender) + relevel(as.factor(ts$ethnic_group), ref = "WHITE") + factor(ever_married), data = ts)
summ(linear1)
#RMSE
sqrt(c(crossprod(linear1$residuals))/length(linear1$residuals))
```

```{r}
linear2 = lm(hosp_time ~ seq_num + icu_los + factor(hospital_expire_flag), data = ts)
summ(linear2)
#RMSE
sqrt(c(crossprod(linear2$residuals))/length(linear2$residuals))
```

Overall the linear model doesn't perform very well for either the large (`model1`) or small (`model2`) models. The most likely reason for this is that motor vehicle accidents are linked to a large variety of trauma-type injuries of different severity; we can see that demographic factors do not factor strongly into either model. It is worth noting that the diagnosis priority had a statistically significant positive regression coefficient, suggesting that multiple complications from a vehicular accident (and thus a lower priority for the diagnosis of the vehicle accident itself) increases length of hospital stay.

---

### Neural Network Regression

```{r, include = F}
library(keras)
library(caret)
library(mltools)
library(data.table)
```

### Create train and test sets

```{r}
preds = ts %>% select(hosp_time, seq_num, icu_los, hospital_expire_flag, age, ethnic_group, ever_married)

fct_relevel(preds$ethnic_group, "WHITE") -> preds$ethnic_group

preds %>% mutate(hospital_expire_flag = hospital_expire_flag, ever_married = as.numeric(ever_married)) -> preds

preds_dummied = one_hot(data.table(na.omit(preds)), cols = c("ethnic_group"))

trainIndex = createDataPartition(preds_dummied$hosp_time, p = 0.7, list = FALSE, times = 1)

train = preds_dummied[trainIndex, ]
test = preds_dummied[-trainIndex, ]

train %>% select(-hosp_time) -> x_train
train %>% select(hosp_time) -> y_train

test %>% select(-hosp_time) -> x_test
test %>% select(hosp_time) -> y_test

train_sc = scale(x_train)
col_means_train = attr(train_sc, "scaled:center")
col_stddevs_train = attr(train_sc, "scaled:scale")
test_sc = scale(x_test, center = col_means_train, scale = col_stddevs_train)

dim(train_sc)
dim(test_sc)
```

### Define simple multi-layer NN model

```{r}
model3 = keras_model_sequential()

model3 %>%
  layer_dense(units = 64, activation = "relu", input_shape = dim(train_sc)[2]) %>%
  layer_dense(units = 64, activation = "relu") %>%
  layer_dense(units = 1, activation = "linear")

summary(model3)
```

```{r}
model3 %>% compile(
  loss = "mse",
  optimizer = optimizer_rmsprop(),
  metrics = c("mean_squared_error")
)

history = model3 %>% fit(as.matrix(train_sc), as.matrix(y_train),
  epochs = 50, batch_size = 50,
  validation_split = 0.2
)
```

```{r}
nnr = plot(history, metrics = "mean_squared_error", smooth = F) +
  xlab("Epoch") +
  ylab("Mean Squared Error") +
  ggtitle("Neural Network Regression Scoring") +
  theme(legend.title = element_blank(),
        legend.text = element_text(c("Training", "Validation")))

nnr

ggsave("nnr.png", plot = nnr)
```

Comparing the RMSE of neural network regression, the training set performs about the same as the standard linear modell; the validation set has a better outcome.
