---
title: "Biostat 203B Homework 3 - Reddit Scraping"
subtitle: "Collecting historical Reddit posts about Coronavirus"
author: "George Dewey"
date: "2/25/2020"
output: html_document
---

This document demonstrates data collection of Reddit posts about the COVID-19 Outbreak.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("reticulate")
use_python('C:/Users/George/Anaconda3/python.exe')
clientid = Sys.getenv("client_id")
clientsec = Sys.getenv("client_secret")
```

Goal: collect the top 50 posts by date from 3 subreddits: `news`, `science`, and `coronavirus`. I believe these 3 communities will present a consistent distribution of discussion topics regarding the COVID-19 Outbreak. The shiny app will randomly select 1 post from each subreddit and display them alongside associated graphs.

Note: PushShiftAPI has some irregularities regarding submission score - to measure post activity I chose the top posts by number of comments.

## Collect data using Python

```{python, eval=FALSE}
import pandas as pd
import praw
from psaw import PushshiftAPI
import datetime as dt

api = PushshiftAPI()

# Choose start date as start of outbreak - virus data started 01/22/2020 
start_time = int(dt.datetime(2020, 1, 22).timestamp())

days_list = [start_time + i*60*60*24 for i in list(range(1, 37))]

# Grab threads from each of the three subreddits
tl_news = []
for i in range(0, 36):
    start_time = days_list[i]
    tl_news.append(list(api.search_submissions(q = 'coronavirus',
                                          after = start_time,
                                          before = start_time + 60 * 60 * 24,
                                          subreddit = 'news',
                                          filter = ["author", "url", "title"], 
                                          sort = 'desc',
                                          sort_type = 'num_comments',
                                          limit = 50)))

tl_sci = []
for i in range(0, 36):
    start_time = days_list[i]
    tl_sci.append(list(api.search_submissions(q = 'coronavirus',
                                          after = start_time,
                                          before = start_time + 60 * 60 * 24,
                                          subreddit = 'science',
                                          filter = ["author", "url", "title"], 
                                          sort = 'desc',
                                          sort_type = 'num_comments',
                                          limit = 50)))

tl_corona = []
for i in range(0, 36):
    start_time = days_list[i]
    tl_corona.append(list(api.search_submissions(after = start_time,
                                          before = start_time + 60 * 60 * 24,
                                          subreddit = 'coronavirus',
                                          filter = ["author", "url", "title"], 
                                          sort = 'desc',
                                          sort_type = 'num_comments',
                                          limit = 50)))

# Stack comments by subreddit into single dataframe and convert to csv

sci_df = pd.DataFrame(tl_sci[0])

for i in range(1, len(tl_sci)):
   sci_df = sci_df.append(pd.DataFrame(tl_sci[i]))

sci_df = sci_df.drop(columns = ['created', 'd_'])

sci_df.to_csv('sci.csv', index=False)

corona_df = pd.DataFrame(tl_corona[0])

for i in range(1, len(tl_corona)):
    corona_df = corona_df.append(pd.DataFrame(tl_corona[i]))

corona_df = corona_df.drop(columns = ['created', 'd_'])

corona_df.to_csv('corona.csv', index=False)

news_df = pd.DataFrame(tl_news[0])

for i in range(1, len(tl_news)):
    news_df = news_df.append(pd.DataFrame(tl_news[i]))

news_df = news_df.drop(columns = ['created', 'd_'])

news_df.to_csv('news.csv', index=False)
```

---

```{r, include = F}
library(tidyverse)
library(lubridate)

sci = read_csv("sci.csv")
corona = read_csv("corona.csv")
news = read_csv("news.csv")
```

## Convert UTC datetimes to date to match user input and export

```{r}
sci = sci %>% mutate(created_date = as_date(as_datetime(created_utc)))
corona = corona %>% mutate(created_date = as_date(as_datetime(created_utc)))
news = news %>% mutate(created_date = as_date(as_datetime(created_utc)))

write_csv(sci, "sci_clean.csv")
write_csv(corona, "corona_clean.csv")
write_csv(news, "news_clean.csv")
```




